Master's Project Journal

### 2019-04-11 thesis/FPGABasedADPLLs.tex
loop filters

\section{Chapter Overview}
The first step in creating an \ac{FPGA} based network of \ac{ADPLL}s is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of an \ac{FPGA} necessitates a number of compromises in the design of a given block which limits transferability to \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in an \ac{FPGA} based \ac{ADPLL} made. A number of blocks have implement purely digital circuitry and as such can be transferred in their entirety from an \ac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail.
\begin{figure}[h] %TODO redo this image
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\ac{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:dist_adpll}
\end{figure}

@@

\section{Frequency Divider}
Many \ac{PLL}s, be they digital or analogue, incorporate a frequency divider in the feedback path so that comparison can be made with a reference operating at a lower frequency. This is a feature that is advantageous for \ac{ADPLL} networks, as comparison using this divided clock allows the synchronisation signal sent between modules to be of a frequency orders of magnitude lower than that of the clock distribution network and thus require circuitry that consumes significantly less energy. This division can be obtained trivially on an \ac{FPGA} by a number of means, of which this thesis will mention just two. The first of these methods allows for integer division, by inverting a signal each time a counter reaches a certain value, $k$. When this condition is true the value stored in the counter is also reset to zero, thus implementing division of the generated clock frequency by $2k$.
\begin{figure}[h]%
	\centering
	\subfloat[Divider 1]{{\includegraphics[width=0.6\textwidth]{../divider1} }}\\
	\subfloat[Divider 2]{{\includegraphics[width=0.6\textwidth]{../divider2} }}%
	\caption{Frequency divider \ac{RTL} diagrams.}
	\label{fig:divs}
\end{figure}
By limiting to division by powers of two, the divider can be simplified with the removal of any comparison logic. One flip-flop and an inverter are required per power of two. As in Figure \ref{fig:divider_2}, the inverted $\overline{Q}$ output of each flip-flop is inverted and connected to the $D$ input, with the non-inverted $Q$ of the previous flip-flop acting as the clock. The first flip-flop in the chain is clocked by the \ac{DCO} output.

@@

\section{Loop Filter}

>>>>

\section{Chapter Overview}
The first step in creating an \ac{FPGA} based network of \ac{ADPLL}s is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of an \ac{FPGA} necessitates a number of compromises in the design of a given block which limits transferability to \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in an \ac{FPGA} based \ac{ADPLL} made. A number of blocks have implement purely digital circuitry and as such can be transferred in their entirety from an \ac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail. A \ac{RTL} diagram of the \ac{ADPLL} topology that will be assumed in the following sections is shown in Figure \ref{fig:network_adpll}.
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\ac{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:network_adpll}
\end{figure}

Important to note is that despite a Cartesian grid leading to four neighbours, only two phase detectors are required per oscillator. Borrowing matrix indexing, and choosing the oscillator located at index $i,j$, $i,j \neq 1,N$ where $N$ is the number of oscillators in each direction, comparison need only be done with the signals generated by oscillators located at $i-1,j$ and $i,j-1$. The comparison with oscillators $i+1,j$ and $i,j+1$ will be carried out in those \ac{ADPLL}s, and the negation of the errors measured used in the \ac{ADPLL} located at $i,j$.

@@

\section{Frequency Divider}
Many \ac{PLL}s, be they digital or analogue, incorporate a frequency divider in the feedback path so that comparison can be made with a reference operating at a lower frequency. This is a feature that is advantageous for \ac{ADPLL} networks, as comparison using this divided clock allows the synchronisation signal sent between modules to be of a frequency orders of magnitude lower than that of the clock distribution network and thus require circuitry that consumes significantly less energy. This division can be obtained trivially on an \ac{FPGA} by a number of means, of which this thesis will mention just two. The first of these methods allows for integer division, by inverting a signal each time a counter reaches a certain value, $k$. When this condition is true the value stored in the counter is also reset to zero, thus implementing division of the generated clock frequency by $2k$.
\begin{figure}[h]%
	\centering
	\subfloat[Divider 1]{{\includegraphics[width=0.6\textwidth]{../divider1} }}\\
	\subfloat[Divider 2]{{\includegraphics[width=0.6\textwidth]{../divider2} }}%
	\caption{Frequency divider \ac{RTL} diagrams.}
	\label{fig:divs}
\end{figure}
By limiting to division by powers of two, the divider can be simplified with the removal of any comparison logic. One flip-flop and an inverter are required per power of two. As in Figure \ref{fig:divs} (b), the inverted $\overline{Q}$ output of each flip-flop is inverted and connected to the $D$ input, with the non-inverted $Q$ of the previous flip-flop acting as the clock. The first flip-flop in the chain is clocked by the \ac{DCO} output.

@@

\section{Loop Filter}
The \acl{LF} in an \ac{ADPLL} can be implemented as a \ac{PI} controller, a circuit that can be implemented in a wide variety of ways, with the exact implementation having little impact on behaviour. The first choice to be made is whether the designer wishes to implement the controller as an \ac{FIR} or \ac{IIR} filter. Both methods will produce a controller that fundamentally behaves the same way, but will have a starkly different implementation. In both cases the filter can be implemented using either fractional or integer representation of the data, depending on the wishes of the designer. Carrying out computations using fractional values may be conceptually easier to understand as the magnitude of individual paths or taps will be independent of another, whereas using integer representations allows for a simpler implementation. If fractional representation is used, it will be done with fixed-point arithmetic as the time complexity of floating point calculations is too great and the potential benefits are inconsequential.

\subsection{\ac{FIR} \acl{LF}}
The \ac{FIR} implementation of the \acl{LF} will take the form presented in Figure \ref{fig:fir_pi} (a), and consists series of time delays, or ``taps'', the output of which is multiplied by a coefficient and the result of each multiplier added together to form the filter output. Various tools, Matlab for example, can be used to compute the coefficients and number of taps required to obtain a specific filter response. Using an \ac{FIR} filter is advantageous as it is always stable due to the lack of a feedback path, however a large number of taps are required to obtain a frequency response akin to that of the \ac{IIR} implementation, thus requiring an equal number of multipliers and $taps-1$ adders. Compared to an \ac{IIR}, which requires just a two adders, two multipliers and a flip-flop this has a much greater hardware footprint. It is possible to condense the \ac{FIR} multiplication and addition to a single multiplier and adder, depicted in \ref{fig:fir_pi} (b), however this will take as many clock cycles to complete as there are taps. This, however, is not viable in this system as the filter must produce its output in a single clock cycle.

The filter can be implemented using fixed-point arithmetic, in which the designer must ensure alignment of the radix in additions and correct radix position in multiplication results. Consequently the fractional width of the each multiplication need not be the same, most commonly used when the proportional gain, the output of the 0th tap multiplication, is significantly larger than that of the other taps. Integer arithmetic instead assumes all calculations are carried out using integers, and output of the addition stage is divided by a large number to form the module output. The integer approach simplifies the task of the designer and reduces the likelihood of incorrect calculations due to misalignment.

\subsection{\ac{IIR} \acl{LF}}
A \ac{PI} controller can be easily implemented as an \ac{IIR} filter in hardware, as in \Figure \ref{fig:iir_pi}, requiring significantly fewer circuit elements than an \ac{FIR} based design as the long series of taps and multipliers is not required, as instead integration is performed by a single multiplier and an accumulator. As before the frequency response of the controller can be calculated using the following equation, where $\alpha$ is the proportional and $\beta$ the integral gain:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}
\end{equation*}

As the scaling factor used in this multiplier is the $k_i$ gain itself, reconfiguration of the integral path of the \acl{LF} can be carried out solely by varying this gain, as opposed to the \ac{FIR} solution where the value of each tap must be recalculated to change the response of the system. The combination of reconfigurability and simpler hardware make the \ac{IIR} the more desirable \ac{LF} design.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi}
	\caption[Basic \ac{IIR} \ac{PI} loop filter]{Basic \ac{IIR} \ac{PI} loop filter.}
	\label{fig:iir_pi}
\end{figure}

Again both integer and fixed-point methods of \ac{LF} design are possible, with the integer method requiring logic combining the two paths to preserve the relationship between the proportional and integral gains, which from the work of Koskin \textit{et al} is known to be several binary orders of magnitude \cite{koskin2018generation}. Failure to preserve a relationship of this order may lead to the filter becoming unstable, as in the upper-left region of Figure \ref{fig:gain_grid}. In the fixed-point system the relationship between $k_p$ and $k_i$ is preserved during the combination of proportional and integral paths as both values have their radix point aligned before addition is carried out.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{example-image-a}%{../koskin_gain_grid} %TODO
	\caption[Stability of \acl{LF} gains]{Stability of \acl{LF} gains \cite{koskin2018generation}.}
	\label{fig:gain_grid}
\end{figure}
